# JOUR405: Statistics for Journalists
## Midterm Exam - Spring 2025

Name: Dylan Schmidt

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences. Start by loading the tidyverse and any other libraries you think you might need.

```{r}
library(tidyverse)
```


## Part 1: Restaurant Health Inspections (15 points)

You want to understand how restaurants in Montgomery County are performing on health inspections. The first dataset contains restaurant health inspection scores for restaurants in Montgomery County. The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible and the compliance score and grade. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv` and complete these tasks:

### Tasks:
1. Calculate the mean and standard deviation of compliance scores (5 points)
2. Create a histogram of the compliance scores with a vertical line showing the mean (5 points)
3. Write 2-3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What would be newsworthy about this distribution? What's the story here? (5 points).

```{r}
inspection_scores <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
inspection_scores |> summarize(mean = mean(compliance_score), sd = sd(compliance_score))
inspection_scores |> 
  ggplot() +
  geom_histogram(aes(x = compliance_score), binwidth = 2) +
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(compliance_score) - sd(compliance_score)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(compliance_score) + sd(compliance_score)), color = "blue", linetype = "dashed", size = 1)
```
The standard deviation and histogram tell us that most of the restaurants' compliance scores are clustered right around the mean of 96. Since the standard deviation is five, the majority of the compliance scores data is clustered between 91 and 100. However, there are a couple outliers with really low scores. I think the biggest story here would be on Kenny's Sub Shop who had a compliance score of 42, well below the next closest place which had a compliance score of 54. 

## Part 2: High School Athletics (25 points)

You are reporting a story about high school sports participation in Maryland and want to see if there are differences between boys and girls. The second dataset shows participation numbers in high school sports across Maryland counties in 2024, broken down by sport and sex. Load the data from: `https://raw.githubusercontent.com/example/md_hs_sports_2024.csv` and complete these tasks:

### Tasks:
1. Calculate the correlation between boys' and girls' participation (5 points)
2. Add two columns called total and girls_pct using mutate(), with the total adding together boys and girls and girls_pct being the percentage of the total represented by girls participants. (5 points)
3. Create a scatterplot showing this relationship, adding a line of best fit (5 points)
4. In 2-3 sentences, explain what the correlation coefficient and scatterplot reveal about equity in Maryland high school sports participation. How do you interpret the school districts that are below the line vs those that are above? Which school districts are most worth examining further, and why? (10 points)

```{r}
sports_participation <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv")
sports_participation |> summarize(correlation = cor(boys, girls, method="pearson"))
total <- sports_participation |> mutate(total = boys + girls)
girls_percentage <- total |> mutate(girls_pct = girls/total)
girls_percentage |> 
  ggplot() +
  geom_point(aes(x=total, y=girls_pct)) +
  geom_smooth(aes(x=total, y=girls_pct), method="lm")
```
A correlation coefficient of 98 and a line of best fit which stays between 0.46 and 0.48 means that most school districts are balanced between girls and boys participation in high school sports. The school districts below the line of best fit have a smaller percentage of girls compared to others with the same total number of athletes , while those above the line of best fit have a larger percentage of girls compared to others with the same total number of athletes. School districts like Somerset County and Baltimore County should be investigated further as they are the biggest outliers.

## Part 3: Public Transit Ridership (20 points)

You are investigating public transit ridership in the Washington, D.C. area and want to understand the patterns of daily bus and rail ridership. The third dataset contains daily bus and rail ridership totals from WMATA for the past year. Load the data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv and do the following:

### Tasks:
1. Calculate the average bus and rail ridership and standard deviation using summarize() (5 points)
2. Using the process you used in class, take a random sample daily ridership numbers and calculate the sample means and deviations for bus and rail. The number in the sample is up to you, but explain why you chose what you did. Compare this to the stats you generated in step 1. (5 points)
3. Using group_by() and summarize(), calculate the means for bus and rail ridership for each weekday. Describe the overall pattern of ridership for bus and rail - which days stand out and why? Are there differences between bus and rail in the standard deviation values? (10 points)


```{r}
transit_ridership <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
transit_ridership |> summarize(mean_bus = mean(bus), sd_bus = sd(bus))
transit_ridership |> summarize(mean_rail = mean(rail), sd_rail = sd(rail))
sample_40 <- transit_ridership |> 
  sample_n(40)
sample_40 |> summarize(sample_mean_bus = mean(bus), sample_sd_bus = sd(bus))
sample_40 |> summarize(sample_mean_rail = mean(rail), sample_sd_rail = sd(rail))
transit_weekday <- transit_ridership |> 
  group_by(weekday)
  summarize(mean_bus_monday = mean(monday)
```



## Part 4: Maryland Car Theft Rates (20 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to analyze the data to find out which counties have the highest rates. The fourth dataset contains car theft statistics for Maryland counties in 2023 and population. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv` and complete the following tasks:

### Tasks:
1. Using mutate, add a column that calculates the rate of car thefts for each county - you need to choose the per capita rate (5 points)
2. Calculate the median car theft rate and the total number of car thefts statewide. Which counties have rates above the median, and what percentage of all car thefts occur in those counties? (5 points)
3. Write 2-3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What's the lede of a story about your findings? (10 points)


```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
rate_car_thefts <- rate_car_thefts |> mutate(car_thefts_2023_rate = `2023`/population * 10000)
median_car_thefts_2023_rate <- rate_car_thefts |> summarize(median = median(car_thefts_2023_rate))
total <-rate_car_thefts|> mutate(total = `2022` + `2023`)
```

## Part 5: Data Analysis Scenario (20 points)

You receive a tip that local emergency response times have gotten significantly worse over the past year. You obtain monthly data on response times for police, fire and ambulance calls.

Write 3-4 sentences (no code!) explaining:
1. What statistical measures would you calculate to verify this claim? (10 points)
I would calculate the percentage change of response times for each individual unit, comparing last year to this year. For example, what was the average police response time last year compared to this year. This would either support or disprove the tip.
2. What visualizations would help readers understand the trends? (5 points)
A histogram would best help readers understand the trend. 
3. What additional context or data would you need to make this a complete story? (5 points)


### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your analytical choices
- Before submitting, clear your environment and run the entire notebook

Good luck!
