---
title: "HW11: Education Level and Survey Weighting"
author: "Your Name Here"
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Introduction

In this assignment, you'll examine how survey weighting affects our understanding of voting intentions across different education levels, using the data we examined in class. You'll calculate both weighted and unweighted statistics, create visualizations, and reflect on the implications for reporting.

## The Data

The nonvoters dataset contains survey responses about voting intentions and behaviors from a national survey. The survey was conducted prior to an election and includes demographic information like education level.

```{r}
# Load the dataset
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")

# Take a quick look at the data structure
glimpse(nonvoters_data)
```

### Key Variables

- `weight`: Survey weight assigned to each respondent
- `Q21`: Voting intention (1 = Yes, 2 = No, 3 = Unsure/Undecided)
- `educ`: Education level (College, Some college, High school or less)

## Task 1: Education Distribution

First, let's examine the distribution of education levels in our sample. Replace "REPLACE_ME" with the correct variable for education level.

```{r}

education_distribution <- nonvoters_data |>
  count(educ) |>
  mutate(percentage = n / sum(n) * 100) |>
  kable(digits = 1, col.names = c("Education Level", "Count", "Percentage (%)"))

education_distribution
```

## Task 2: Reflection Question

Why might education levels in survey samples often differ from the general population? What factors might cause certain education groups to be over or underrepresented?

People with an education level of high school or less may feel embarrassed to include that information in the survey and therefore may be less inclined to actually answer the survey as a whole. The method that the survey is conducted can also factor into a certain education groups' response rate, as people with high school or less education may not have the same access to technology or the internet. The wording of a question can also deter people away from answering the survey if it uses complex words or jargon they don't understand.


## Task 3: Unweighted Analysis by Education

Now, let's calculate unweighted voting intentions by education level. This is what we would report if we didn't apply any weighting to our sample.

```{r}
# Calculate unweighted voting intentions by education
unweighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Count responses
  summarize(count = n(), .groups = "drop_last") |>
  # Calculate percentages
  mutate(total = sum(count),
         percentage = count / total * 100) |>
  ungroup()

# Create a more readable format with voting intentions as columns
unweighted_educ_summary <- unweighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

kable(unweighted_educ_summary, digits = 1, caption = "Unweighted Voting Intentions by Education Level")
```

## Task 4: Reflection Question

Based just on this unweighted analysis, what headline might you write for a news story about education and voting intentions?

People with an education of high school or less are less likely to vote 

## Task 5: Weighted Analysis by Education

Next, let's apply survey weights to see how this changes our results. Instead of just counting responses, we'll sum the weights for each group. Replace "REPLACE_ME" with the appropriate weight variable

```{r weighted-by-education}

weighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Sum the weights instead of counting
  summarize(weighted_count = sum(weight), .groups = "drop_last") |>
  # Calculate weighted percentages
  mutate(weighted_total = sum(weighted_count),
         weighted_percentage = weighted_count / weighted_total * 100) |>
  ungroup()

# Create a more readable format
weighted_educ_summary <- weighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = weighted_percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

kable(weighted_educ_summary, digits = 1, caption = "Weighted Voting Intentions by Education Level")
```

## Task 6: Reflection Questions

1. How did the percentages change after applying weights? Which education group showed the biggest changes?

The voting intentions of those people with an education of college or some college barely saw any changes. However, the education group "High school or less" saw the biggest change. The percentage of voters who intend to vote decreased by three percent, while the percentage of voters who don't intend to vote or are unsure decreased by just over one percent. 

2. Why might the weighted results be considered more accurate than the unweighted results?

The weighted results may be considered more accurate than the unweighted results because it adjusts certain groups who are over or underrepresented to make it more reflective of the general population.
## Task 7: Comparison of Weighted vs. Unweighted Results

Let's create a direct comparison table to see the differences more clearly.

```{r}

comparison <- unweighted_educ_summary |>
  inner_join(weighted_educ_summary, by = "educ", suffix = c("_unweighted", "_weighted")) |>
  mutate(
    # Calculate the differences between weighted and unweighted percentages
    yes_diff = `Yes (%)_weighted` - `Yes (%)_unweighted`,
    no_diff = `No (%)_weighted` - `No (%)_unweighted`,
    unsure_diff = `Unsure (%)_weighted` - `Unsure (%)_unweighted`
  ) |>
  # Select just the columns we want to display
  select(educ, yes_diff, no_diff, unsure_diff) |>
  rename(
    "Education Level" = educ,
    "Yes (% point diff)" = yes_diff,
    "No (% point diff)" = no_diff,
    "Unsure (% point diff)" = unsure_diff
  )

kable(comparison, digits = 1, caption = "Difference Between Weighted and Unweighted Results (percentage points)")
```

## Task 8: Reflection Question

Which education group shows the largest differences between weighted and unweighted results?

The education group "High school or less" shows the largest difference between weighted and unweighted results.

## Task 9: Visualization

Visualizations can help us see the differences more clearly. Let's create a bar chart comparing weighted and unweighted "Yes" responses by education level. Replace "REPLACE_ME" with the correct variable name

```{r}
educ_viz_data <- bind_rows(
  # Unweighted data
  unweighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses (Q21=1)
    mutate(Type = "Unweighted") |>
    select(Type, educ, percentage),
  
  # Weighted data - 
  weighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses
    mutate(
      Type = "Weighted",
      percentage = weighted_percentage
    ) |>
    select(Type, educ, percentage)
)

# Create a grouped bar chart
ggplot(educ_viz_data, 
       aes(x = educ, y = percentage, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(
    title = "Weighted vs. Unweighted 'Yes' Responses by Education",
    subtitle = "Q21: Do you plan to vote in the November election?",
    y = "Percentage (%)",
    x = "Education Level"
  ) +
  scale_fill_manual(values = c("Unweighted" = "#619CFF", "Weighted" = "#F8766D")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Task 10: Reflection Questions

Does the visualization make it easier to see the differences between weighted and unweighted results? Why or why not?

Yes, the visualization allows us to compare the different percentages side-by-side between weighted and unweighted in order to decipher which one the weighted change had the greatest impact on.
## Task 11: Summary

Based on your analysis of weighted and unweighted results by education level, write a brief (2-3 paragraph) journalistic summary of what you found. You should try to address:

1. How education level relates to voting intentions
2. How weighting affected your understanding of this relationship
3. What this means for interpreting polling results in news reporting

After comparing over 5,000 survey responses, we have determined that a person's education level greatly affects their intentions to vote. Only 73.5% of people with an education level with an education level of high school or less have the intention of voting in the November election. On the other side, 93.4% and 85.7% pf people with college or some college education have the intention of voting in the November election, a steep increase compared to those without.

Weighting also played a factor in the final results of the survey, affecting those same numbers to some extent. The percentage of people who intend to vote in the November election with a college education increased by 0.1, while the percentage within the group of people with some college education decreased by 0.7. The biggest change, though, came from the grouping of people with an education level of high school or less, as it plummeted down 2.7 percent.

This means that polling results are hard to interpret in news reporting due to some groups being over or underrepresented in the poll. Just like how the results of people who intended to vote in the November election changed, weighting plays a role in the accuracy of those polls, making it more representive of the population. Polling results can only be relied upon when they are weighted, and even then, I would shy away from basing your expectations of the results on upcoming elections from the polling results.
## Task 12: Final Reflection Questions

1. Why is it important for journalists to understand survey weighting when reporting on polls?

Understanding survey weighting will decrease the risk of journalists misinterpreting or misrepresenting the findings of polls. It will also prevent journalists from generalizing the results of unweighted polls and assume it is accurate of the general population.

2. How might the differences between weighted and unweighted results affect how you would report on this data?

The differences between weighted and unweighted results will deter me away from generalizing the results of the survey if I were to report on this data. If I were to write a story on these results, I would include both the weighted and unweighted results while also explaining the weighting proccess applied to the initial survey results. I would also explicitly state that the results of the survey are by no means accurate of the whole population and can be just used to determine the potential outcome of the election. 

3. What additional information would you want to know about how the weights were calculated before using this data in a news story?

I would want to know the exact weighting process used so that I could specify it in the story and determine how I should decipher these results when writing my story.
