---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE
I noticed that a couple of the workers have salaries of zero. However, the thing that stuck out to me most was that the majority of these workers have an employment status of detailee and not employee. But some of the workers that are labeled as detailee's have a salary, while a couple employees have no salary.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean = salary/564)
wh_salaries |> summarize(median = median(salary))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
mean_salary <- wh_salaries |> summarize(mean = mean(salary))
median_salary <- wh_salaries |> summarize(median = median(salary))
min_salary <- wh_salaries |> reframe(min = min(salary))
max_salary <- wh_salaries |>  reframe(max = max(salary))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE
The results of the mean and median salaries tell me that the data is skewed to the left and that there are values on the lower end of the scale dragging the mean number down. The workers with a salary of zero are dragging the mean below the median. The minimum salary tells me that policy advisors, senior policy advisors, senior directors, and directors who are getting paid nothing. However, the associate directors are getting paid way more. The maximum salary is well above the next closest salary — nearly $60,000 — is getting paid a good bit more than others with the same job title.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE
My best story idea would be investigating why certain government employees are unpaid. The employees are getting paid nothing for their work, while the next closest cluster of workers is getting paid 20 thousand dollars.
