---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
candy <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.

```{r}
sample_5 <- candy |> 
  sample_n(5)

sample_10 <- candy |> 
  sample_n(10)
```

### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.

```{r}
sample_5 |> summarize(mean = mean(black), sd = sd(black))
sample_10 |> summarize(mean = mean(black), sd = sd(black))
```

### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4
```{r}
candy |> summarize(mean = mean(black), sd = sd(black))
```

### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?
The sample mean of the five rows is closer to the population mean, 73.8 to 73.28. The sample mean of 10 rows is way off that of the population mean. This is similar with the standard deviation, as the sample of five rows is closer to the population value than the sample of 10 rows.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

This tells me that the act of random sampling will see different results every time. If I took a sample of five rows of this same data multiple times, I would see different means and standard deviations. This is the same with different sample sizes as the results will consistently vary. In regards to the M&Ms dataset, it tells that a sample of five rows would more accurately represent the population than a sample of 10 rows.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
